\textbf{Evaluation of machine learning methods for 16S rRNA gene data}

### 2019-07-12 submission/manuscript.Rmd
fix makefile
update

We reported the testing AUROC values over 100 randomizations of the initial data-split. The results showed that depending on the data-split, the testing AUROC value showed great variability [Figure 2]. The testing AUROC values within each model varied `r paste(round(mean((all %>% group_by(model) %>% summarise(difference=max(test_aucs)-min(test_aucs)))$difference), 2))` on average across the seven models. For instance, the lowest AUROC value of the random forest model was `r paste(round(min((all %>% filter(model=="Random_Forest"))$test_aucs), 2))` whereas the highest was `r paste(round(max((all %>% filter(model=="Random_Forest"))$test_aucs), 2))`.

@@

We explained the feature importances in non-linear models; SVM with radial basis kernel, decision tree, random forest and XGBoost, using a metod called permutation importance on the held-out test data.  Permutation importance analysis is where we randomly permute non-correlated features individually and groups of highly correlated features together and calculate how much the predictive performance of the model (i.e AUROC values) decrease if those specific OTUs or group of OTUs are permuted randomly. We ranked the OTUs based on how much they decreased the test AUROC values if they were randomly permuted, with the largest decrease ranking highest. The top 5 OTUs with the largest negative impact on testing AUROC overlapped in tree-based models [Figure 4]. Specifically, permuting *Peptostreptococcus* (OTU00367) abundances randomly, dropped the predictive performances the most in all tree-based methods [Figure S3]. Decision tree, random forest and XGBoost models' predictive performance dropped from `r paste(round(dt_median, 2))` base testing median AUROC to `r paste(round(median((dt_imp %>% filter(names=="Otu00367"))$new_auc), 2))`, from `r paste(round(rf_median, 2))` to `r paste(round(median((rf_imp %>% filter(names=="Otu00367"))$new_auc), 2))` and from `r paste(round(xgboost_median, 2))` to `r paste(round(median((xgboost_imp %>% filter(names=="Otu00367"))$new_auc), 2))`, respectively [Figure 4].

To highlight the differences between the two interpretation methods, we used permutation importance to interpret linear models as well [Figure S3]. The importance rankings of OTUs that discriminate between having an SRN or not, changed between different interpretation methods. L1-regularized SVM with linear kernel picked out some of the same OTUs (OTU00822, OTU01239, OTU00609) as important in making a prediction in the two interpretation methods; feature rankings based on weights [Figure 3] and permutation importance [Figure S3]. Similarly,  L2-regularized SVM and L2-regularized logistic regression picked out some of the same OTUs in both interpretation models, OTU00659 and OTU00012, respectively. However, for all the linear models, the rankings of these features were different due to the collinearity in microbial communities.

@@

**Figure 2. Generalization and classification performance of ML models using AUROC values of all cross validation and testing performances. **   The median AUROC for diagnosing individuals with SRN using bacterial abundances was higher than chance (depicted by horizontal line at 0.50) for all the ML models. Discriminative performance of random forest model was higher than other ML models. The boxplot shows quartiles at the box ends and the statistical median as the horizontal line in the box. The whiskers show the farthest points that are not outliers. Outliers are data points that are not within 3/2 times the interquartile ranges. 
Abbreviations: SRN, screen-relevant neoplasias; AUROC, area under the receiver operating characteristic curve; SVM, support vector machine; XGBoost, extreme gradient boosting
\newpage
\includegraphics[height=15cm, width=10cm]{Figure_3.png}

@@

\newpage
\includegraphics{Figure_S2.png}

**Figure S2. Hyperparameter setting performances for non-linear models.** (A) Decision tree (B) Random forest (C) SVM with radial basis kernel (D) XGoost mean cross-validation AUROC values when different hyperparameters are used in training the model. The differences in AUROC values when hyperparameters change show that hyperparameter tuning is a crucial step in building a ML model.
\newpage
\includegraphics[height=17.5cm, width=13cm]{Figure_S4.png}

>>>>

We reported the testing AUROC values over 100 randomizations of the initial data-split. The results showed that depending on the data-split, the testing AUROC values showed great variability [Figure 2]. The testing AUROC values within each model varied `r paste(round(mean((all %>% group_by(model) %>% summarise(difference=max(test_aucs)-min(test_aucs)))$difference), 2))` on average across the seven models. For instance, the lowest AUROC value of the random forest model was `r paste(round(min((all %>% filter(model=="Random_Forest"))$test_aucs), 2))` whereas the highest was `r paste(round(max((all %>% filter(model=="Random_Forest"))$test_aucs), 2))`.

@@

We explained the feature importances in non-linear models; SVM with radial basis kernel, decision tree, random forest and XGBoost, using a method called permutation importance on the held-out test data.  Permutation importance analysis is where we randomly permute non-correlated features individually and groups of highly correlated features together. We then calculate how much the predictive performance of the model (i.e AUROC values) decrease when each OTU or group of OTUs is permuted randomly. We ranked the OTUs based on how much they decreased the median testing AUROC; the OTU with the largest decrease ranking highest. The top 5 OTUs with the largest negative impact on testing AUROC overlapped in tree-based models [Figure 4]. Specifically, permuting *Peptostreptococcus* (OTU00367) abundances randomly, dropped the predictive performances the most in all tree-based methods [Figure 4]. Decision tree, random forest and XGBoost models' predictive performance dropped from `r paste(round(dt_median, 2))` median AUROC to `r paste(round(median((dt_imp %>% filter(names=="Otu00367"))$new_auc), 2))`, from `r paste(round(rf_median, 2))` to `r paste(round(median((rf_imp %>% filter(names=="Otu00367"))$new_auc), 2))` and from `r paste(round(xgboost_median, 2))` to `r paste(round(median((xgboost_imp %>% filter(names=="Otu00367"))$new_auc), 2))`, respectively [Figure 4].

To highlight the differences between the two interpretation methods, we used permutation importance to interpret linear models as well [Figure S3]. L1-regularized SVM with linear kernel picked out some of the same OTUs (OTU00822, OTU01239, OTU00609) as important in the two interpretation methods; feature rankings based on weights [Figure 3] and permutation importance [Figure S3]. Similarly,  L2-regularized SVM and L2-regularized logistic regression picked out some of the same OTUs in both interpretation methods, OTU00659 and OTU00012, respectively. However, for all the linear models, the rankings of these features were different due to the collinearity in microbial communities.

@@

**Figure 2. Generalization and classification performance of ML models using AUROC values of all cross validation and testing performances. **   The median AUROC for diagnosing individuals with SRN using bacterial abundances was higher than chance (depicted by horizontal line at 0.50) for all the ML models. Discriminative performance of random forest model was higher than other ML models. The boxplot shows quartiles at the box ends and the statistical median as the horizontal line in the box. The whiskers show the farthest points that are not outliers. Outliers are data points that are not within 3/2 times the interquartile ranges. 
Abbreviations: SRN, screen-relevant neoplasias; AUROC, area under the receiver operating characteristic curve; SVM, support vector machine; XGBoost, extreme gradient boosting
\newpage
\includegraphics[height=17cm, width=12cm]{Figure_3.png}

@@

\newpage
\includegraphics[height=30cm, width=15cm]{Figure_S2.png}

**Figure S2. Hyperparameter setting performances for non-linear models.** (A) Decision tree (B) Random forest (C) SVM with radial basis kernel (D) XGoost mean cross-validation AUROC values when different hyperparameters are used in training the model. The differences in AUROC values when hyperparameters change show that hyperparameter tuning is a crucial step in building a ML model.
\newpage
\includegraphics[height=17.5cm, width=13cm]{Figure_S3.png}

### 2019-07-12 submission/manuscript.tex
add comment on importance
fix makefile
update

\subsection{Importance}\label{importance}

@@

We reported the testing AUROC values over 100 randomizations of the
initial data-split. The results showed that depending on the data-split,
the testing AUROC value showed great variability {[}Figure 2{]}. The
testing AUROC values within each model varied 0.23 on average across the
seven models. For instance, the lowest AUROC value of the random forest
model was 0.59 whereas the highest was 0.81.

@@

We explained the feature importances in non-linear models; SVM with
radial basis kernel, decision tree, random forest and XGBoost, using a
metod called permutation importance on the held-out test data.
Permutation importance analysis is where we randomly permute
non-correlated features individually and groups of highly correlated
features together and calculate how much the predictive performance of
the model (i.e AUROC values) decrease if those specific OTUs or group of
OTUs are permuted randomly. We ranked the OTUs based on how much they
decreased the test AUROC values if they were randomly permuted, with the
largest decrease ranking highest. The top 5 OTUs with the largest
negative impact on testing AUROC overlapped in tree-based models
{[}Figure 4{]}. Specifically, permuting \emph{Peptostreptococcus}
(OTU00367) abundances randomly, dropped the predictive performances the
most in all tree-based methods {[}Figure S3{]}. Decision tree, random
forest and XGBoost models' predictive performance dropped from 0.6 base
testing median AUROC to 0.52, from 0.69 to 0.68 and from 0.68 to 0.65,
respectively {[}Figure 4{]}.

To highlight the differences between the two interpretation methods, we
used permutation importance to interpret linear models as well {[}Figure
S3{]}. The importance rankings of OTUs that discriminate between having
an SRN or not, changed between different interpretation methods.
L1-regularized SVM with linear kernel picked out some of the same OTUs
(OTU00822, OTU01239, OTU00609) as important in making a prediction in
the two interpretation methods; feature rankings based on weights
{[}Figure 3{]} and permutation importance {[}Figure S3{]}. Similarly,
L2-regularized SVM and L2-regularized logistic regression picked out
some of the same OTUs in both interpretation models, OTU00659 and
OTU00012, respectively. However, for all the linear models, the rankings
of these features were different due to the collinearity in microbial
communities.

@@

\textbf{Figure 2. Generalization and classification performance of ML
models using AUROC values of all cross validation and testing
performances. } The median AUROC for diagnosing individuals with SRN
using bacterial abundances was higher than chance (depicted by
horizontal line at 0.50) for all the ML models. Discriminative
performance of random forest model was higher than other ML models. The
boxplot shows quartiles at the box ends and the statistical median as
the horizontal line in the box. The whiskers show the farthest points
that are not outliers. Outliers are data points that are not within 3/2
times the interquartile ranges. Abbreviations: SRN, screen-relevant
neoplasias; AUROC, area under the receiver operating characteristic
curve; SVM, support vector machine; XGBoost, extreme gradient boosting
\newpage
\includegraphics[height=15cm, width=10cm]{Figure_3.png}

@@

\includegraphics{Figure_S2.png}

\textbf{Figure S2. Hyperparameter setting performances for non-linear
models.} (A) Decision tree (B) Random forest (C) SVM with radial basis
kernel (D) XGoost mean cross-validation AUROC values when different
hyperparameters are used in training the model. The differences in AUROC
values when hyperparameters change show that hyperparameter tuning is a
crucial step in building a ML model. \newpage
\includegraphics[height=17.5cm, width=13cm]{Figure_S4.png}

>>>>

\subsection{Importance (needs work)}\label{importance-needs-work}

@@

We reported the testing AUROC values over 100 randomizations of the
initial data-split. The results showed that depending on the data-split,
the testing AUROC values showed great variability {[}Figure 2{]}. The
testing AUROC values within each model varied 0.23 on average across the
seven models. For instance, the lowest AUROC value of the random forest
model was 0.59 whereas the highest was 0.81.

@@

We explained the feature importances in non-linear models; SVM with
radial basis kernel, decision tree, random forest and XGBoost, using a
method called permutation importance on the held-out test data.
Permutation importance analysis is where we randomly permute
non-correlated features individually and groups of highly correlated
features together. We then calculate how much the predictive performance
of the model (i.e AUROC values) decrease when each OTU or group of OTUs
is permuted randomly. We ranked the OTUs based on how much they
decreased the median testing AUROC; the OTU with the largest decrease
ranking highest. The top 5 OTUs with the largest negative impact on
testing AUROC overlapped in tree-based models {[}Figure 4{]}.
Specifically, permuting \emph{Peptostreptococcus} (OTU00367) abundances
randomly, dropped the predictive performances the most in all tree-based
methods {[}Figure 4{]}. Decision tree, random forest and XGBoost models'
predictive performance dropped from 0.6 median AUROC to 0.52, from 0.69
to 0.68 and from 0.68 to 0.65, respectively {[}Figure 4{]}.

To highlight the differences between the two interpretation methods, we
used permutation importance to interpret linear models as well {[}Figure
S3{]}. L1-regularized SVM with linear kernel picked out some of the same
OTUs (OTU00822, OTU01239, OTU00609) as important in the two
interpretation methods; feature rankings based on weights {[}Figure 3{]}
and permutation importance {[}Figure S3{]}. Similarly, L2-regularized
SVM and L2-regularized logistic regression picked out some of the same
OTUs in both interpretation methods, OTU00659 and OTU00012,
respectively. However, for all the linear models, the rankings of these
features were different due to the collinearity in microbial
communities.

@@

\textbf{Figure 2. Generalization and classification performance of ML
models using AUROC values of all cross validation and testing
performances. } The median AUROC for diagnosing individuals with SRN
using bacterial abundances was higher than chance (depicted by
horizontal line at 0.50) for all the ML models. Discriminative
performance of random forest model was higher than other ML models. The
boxplot shows quartiles at the box ends and the statistical median as
the horizontal line in the box. The whiskers show the farthest points
that are not outliers. Outliers are data points that are not within 3/2
times the interquartile ranges. Abbreviations: SRN, screen-relevant
neoplasias; AUROC, area under the receiver operating characteristic
curve; SVM, support vector machine; XGBoost, extreme gradient boosting
\newpage
\includegraphics[height=17cm, width=12cm]{Figure_3.png}

@@

\includegraphics[height=30cm, width=15cm]{Figure_S2.png}

\textbf{Figure S2. Hyperparameter setting performances for non-linear
models.} (A) Decision tree (B) Random forest (C) SVM with radial basis
kernel (D) XGoost mean cross-validation AUROC values when different
hyperparameters are used in training the model. The differences in AUROC
values when hyperparameters change show that hyperparameter tuning is a
crucial step in building a ML model. \newpage
\includegraphics[height=17.5cm, width=13cm]{Figure_S3.png}
